---
title: "Causal inference with a latent IRT confounder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal inference with a latent IRT confounder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

## The model

`bairrtt` fits a causal model in which a latent person trait `theta` is both

* the ability parameter in a two-parameter logistic (2PL) item-response model,
  measured by a matrix of 0/1 item responses, and
* a confounder between a binary treatment `z` and a continuous outcome `y`.

Because `theta` is unobserved and drives both treatment assignment and the
outcome, ignoring it biases the treatment effect. Two BART surfaces (from
[dbarts](https://cran.r-project.org/package=dbarts)) model the outcome
`y ~ f(z, theta)` and the assignment `z ~ f(theta)` flexibly, while the IRT
measurement model ties `theta` to the item responses. The pieces are sampled in
a Gibbs loop:

* `theta` -- one per-person random-walk Metropolis move, then installed into
  both BART models by a joint per-observation sweep that rejects any value which
  would empty a tree leaf (a zero-prior tree);
* the trees -- one BART update each, conditional on `theta`;
* the IRT item parameters `(alpha, beta)` -- drawn conditional on `theta` by the
  gradient-based [WALNUTS](https://github.com/bob-carpenter/walnuts) sampler,
  which adapts during burn-in and is then frozen.

## Simulate data

`simulate_irt_causal()` draws from this model. Treatment depends on `theta`, so
a naive treatment/outcome comparison is confounded.

```{r simulate}
library(bairrtt)

sim <- simulate_irt_causal(n_persons = 300, n_items = 40, ate = -0.2, seed = 42)
str(sim, max.level = 1)

# naive (confounded) difference in means -- biased for the true effect (-0.2)
mean(sim$y[sim$z == 1]) - mean(sim$y[sim$z == 0])
```

## Fit

`irt_causal_bart()` runs the whole sampler. The run below is deliberately small
so the vignette builds quickly; recovery sharpens with more persons/items and
longer chains.

```{r fit}
fit <- irt_causal_bart(
  sim$responses, sim$y, sim$z,
  n_burnin = 100, n_sampling = 250, seed = 42
)
```

## Treatment effect

The estimand is `E[f(1, theta) - f(0, theta)]`, accumulated from the response
surface each scan. Its posterior recovers the true effect (`-0.2`), unlike the
naive comparison above.

```{r ate}
c(mean = mean(fit$ate),
  lower = quantile(fit$ate, 0.025, names = FALSE),
  upper = quantile(fit$ate, 0.975, names = FALSE))

hist(fit$ate, breaks = 30, main = "Posterior of the treatment effect",
     xlab = "ATE")
abline(v = sim$ate, col = "red", lwd = 2)
```

## Item parameters

The IRT item parameters are recovered too. Difficulty is read on the observable
scale --- each item's expected proportion-correct in the population, its
*endorsement* probability --- rather than as the raw latent `beta`, whose scale
is fixed only by convention and is stretched by the finite number of items
measuring each person; discrimination sharpens as that number grows.

```{r items}
alpha_hat <- colMeans(fit$alpha)

# standard-normal population, for integrating each item's characteristic curve
grid <- qnorm(ppoints(200))

# an item's expected proportion-correct over the population: an observable-scale
# difficulty, free of the latent metric that inflates the raw beta coordinate
endorsement <- function(alpha, beta)
  vapply(seq_along(beta),
         function(i) mean(plogis(alpha[i] * (grid - beta[i]))),
         numeric(1))

pi_true <- endorsement(sim$alpha, sim$beta)
pi_hat  <- rowMeans(vapply(seq_len(nrow(fit$beta)),
                           function(s) endorsement(fit$alpha[s, ], fit$beta[s, ]),
                           numeric(ncol(fit$beta))))

c(alpha_cor      = cor(alpha_hat, sim$alpha),
  difficulty_cor = cor(pi_hat,    pi_true))

plot(qlogis(pi_true), qlogis(pi_hat),
     xlab = "true difficulty (logit endorsement)", ylab = "posterior mean",
     main = "Item difficulties"); abline(0, 1, col = "red")
```

## Diagnostics

The per-person `theta` acceptance rate is adapted toward its target (0.44 by
default) during burn-in and then frozen, and the WALNUTS item sampler is frozen
for the sampling phase.

```{r diagnostics}
mean(tail(fit$theta_accept, 250))   # sampling-phase theta acceptance
fit$theta_sd                        # frozen proposal SD
fit$tuning$frozen                   # WALNUTS frozen?
```

## More than one latent trait

Several traits can confound the same treatment and outcome, each measured by its
own item bank. Pass `responses` as a *list* of response matrices, one per trait;
every item loads on exactly one trait, and the banks may differ in item count so
long as they cover the same persons in the same order. Conditional on the traits
the banks factorize exactly, so each gets its own WALNUTS sampler.

The traits are visited in a systematic scan --- trait `k` is proposed and
installed for every person before trait `k + 1` --- because the joint install
into both BART models takes one shared column at a time. That is an ordinary
Metropolis-within-Gibbs sweep; what it does not do is move a person's whole
trait vector as a single accept/reject unit.

```{r simulate2}
sim2 <- simulate_irt_causal(
  n_persons = 300, n_items = c(30, 40), n_traits = 2,
  ate = -0.2, prognostic = c(1.2, 0.8), seed = 42
)
dim(sim2$theta)                       # one column per trait
sapply(sim2$responses, dim)           # 300 x 30 and 300 x 40

fit2 <- irt_causal_bart(
  sim2$responses, sim2$y, sim2$z,
  n_burnin = 100, n_sampling = 250, seed = 42
)
```

The estimand is unchanged in form --- `E[f(1, theta) - f(0, theta)]`, now
conditioning on the whole trait matrix. The per-bank quantities come back as
lists, one entry per trait; everything else keeps its single-trait shape.

```{r ate2}
c(mean = mean(fit2$ate),
  lower = quantile(fit2$ate, 0.025, names = FALSE),
  upper = quantile(fit2$ate, 0.975, names = FALSE))

# item difficulties still track the truth within each bank
vapply(1:2, function(k) cor(colMeans(fit2$beta[[k]]), sim2$beta[[k]]),
       numeric(1))
```

Only the disjoint-bank case is supported. Genuine multidimensional IRT, where
one item loads on several traits, is identified only up to a rotation of the
loadings and needs a constraint that this interface does not express.

## Reusing the WALNUTS item sampler

The item-parameter sampler is also exported on its own, for use inside other
models. Create it once, adapt with `irt_warmup()` while conditioning on a
(possibly changing) `theta`, `irt_freeze()`, then draw with `irt_draw()`.

```{r engine}
s <- irt_item_sampler(sim$responses, alpha = rep(1, 40), beta = rep(0, 40),
                      seed = 1)
for (i in 1:50) invisible(irt_warmup(s, sim$theta))   # adapt
irt_freeze(s)
draw <- irt_draw(s, sim$theta)                         # c(alpha, beta)
head(draw)
```
