---
title: "Causal inference with a latent IRT confounder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal inference with a latent IRT confounder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

## The model

`bairrtt` fits a causal model in which a latent person trait `theta` is both

* the ability parameter in a two-parameter logistic (2PL) item-response model,
  measured by a matrix of 0/1 item responses, and
* a confounder between a binary treatment `z` and a continuous outcome `y`.

Because `theta` is unobserved and drives both treatment assignment and the
outcome, ignoring it biases the treatment effect. Two BART surfaces (from
[dbarts](https://cran.r-project.org/package=dbarts)) model the outcome
`y ~ f(z, theta)` and the assignment `z ~ f(theta)` flexibly, while the IRT
measurement model ties `theta` to the item responses. The pieces are sampled in
a Gibbs loop:

* `theta` -- one per-person random-walk Metropolis move, then installed into
  both BART models by a joint per-observation sweep that rejects any value which
  would empty a tree leaf (a zero-prior tree);
* the trees -- one BART update each, conditional on `theta`;
* the IRT item parameters `(alpha, beta)` -- drawn conditional on `theta` by the
  gradient-based [WALNUTS](https://github.com/bob-carpenter/walnuts) sampler,
  which adapts during burn-in and is then frozen.

## Simulate data

`simulate_irt_causal()` draws from this model. Treatment depends on `theta`, so
a naive treatment/outcome comparison is confounded.

```{r simulate}
library(bairrtt)

sim <- simulate_irt_causal(n_persons = 300, n_items = 40, ate = -0.2, seed = 42)
str(sim, max.level = 1)

# naive (confounded) difference in means -- biased for the true effect (-0.2)
mean(sim$y[sim$z == 1]) - mean(sim$y[sim$z == 0])
```

## Fit

`irt_causal_bart()` runs the whole sampler. The run below is deliberately small
so the vignette builds quickly; recovery sharpens with more persons/items and
longer chains.

```{r fit}
fit <- irt_causal_bart(
  sim$responses, sim$y, sim$z,
  n_burnin = 100, n_sampling = 250, seed = 42
)
```

## Treatment effect

The estimand is `E[f(1, theta) - f(0, theta)]`, accumulated from the response
surface each scan. Its posterior recovers the true effect (`-0.2`), unlike the
naive comparison above.

```{r ate}
c(mean = mean(fit$ate),
  lower = quantile(fit$ate, 0.025, names = FALSE),
  upper = quantile(fit$ate, 0.975, names = FALSE))

hist(fit$ate, breaks = 30, main = "Posterior of the treatment effect",
     xlab = "ATE")
abline(v = sim$ate, col = "red", lwd = 2)
```

## Item parameters

The IRT item parameters are recovered too. Difficulty is read on the observable
scale --- each item's expected proportion-correct in the population, its
*endorsement* probability --- rather than as the raw latent `beta`, whose scale
is fixed only by convention and is stretched by the finite number of items
measuring each person; discrimination sharpens as that number grows.

```{r items}
alpha_hat <- colMeans(fit$alpha)

# standard-normal population, for integrating each item's characteristic curve
grid <- qnorm(ppoints(200))

# an item's expected proportion-correct over the population: an observable-scale
# difficulty, free of the latent metric that inflates the raw beta coordinate
endorsement <- function(alpha, beta)
  vapply(seq_along(beta),
         function(i) mean(plogis(alpha[i] * (grid - beta[i]))),
         numeric(1))

pi_true <- endorsement(sim$alpha, sim$beta)
pi_hat  <- rowMeans(vapply(seq_len(nrow(fit$beta)),
                           function(s) endorsement(fit$alpha[s, ], fit$beta[s, ]),
                           numeric(ncol(fit$beta))))

c(alpha_cor      = cor(alpha_hat, sim$alpha),
  difficulty_cor = cor(pi_hat,    pi_true))

plot(qlogis(pi_true), qlogis(pi_hat),
     xlab = "true difficulty (logit endorsement)", ylab = "posterior mean",
     main = "Item difficulties"); abline(0, 1, col = "red")
```

## Diagnostics

The per-person `theta` acceptance rate is adapted toward its target (0.44 by
default) during burn-in and then frozen, and the WALNUTS item sampler is frozen
for the sampling phase.

```{r diagnostics}
mean(tail(fit$theta_accept, 250))   # sampling-phase theta acceptance
fit$theta_sd                        # frozen proposal SD
fit$tuning$frozen                   # WALNUTS frozen?
```

## Reusing the WALNUTS item sampler

The item-parameter sampler is also exported on its own, for use inside other
models. Create it once, adapt with `irt_warmup()` while conditioning on a
(possibly changing) `theta`, `irt_freeze()`, then draw with `irt_draw()`.

```{r engine}
s <- irt_item_sampler(sim$responses, alpha = rep(1, 40), beta = rep(0, 40),
                      seed = 1)
for (i in 1:50) invisible(irt_warmup(s, sim$theta))   # adapt
irt_freeze(s)
draw <- irt_draw(s, sim$theta)                         # c(alpha, beta)
head(draw)
```
