---
title: "Causal inference with a latent IRT confounder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal inference with a latent IRT confounder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

## The model

`bairrtt` fits a causal model in which a latent person trait `theta` is both

* the ability parameter in a two-parameter logistic (2PL) item-response model,
  measured by a matrix of 0/1 item responses, and
* a confounder between a binary treatment `z` and a continuous outcome `y`.

Because `theta` is unobserved and drives both treatment assignment and the
outcome, ignoring it biases the treatment effect. Two BART surfaces (from
[dbarts](https://cran.r-project.org/package=dbarts)) model the outcome
`y ~ f(z, theta)` and the assignment `z ~ f(theta)` flexibly, while the IRT
measurement model ties `theta` to the item responses. The pieces are sampled in
a Gibbs loop:

* `theta` -- one per-person random-walk Metropolis move, then installed into
  both BART models by a joint per-observation sweep that rejects any value which
  would empty a tree leaf (a zero-prior tree);
* the trees -- one BART update each, conditional on `theta`;
* the IRT item parameters `(alpha, beta)` -- drawn conditional on `theta` by the
  gradient-based [WALNUTS](https://github.com/bob-carpenter/walnuts) sampler,
  which adapts during burn-in and is then frozen.

## What has to be true

`bairrtt` returns a causal effect only under assumptions it cannot check. They
are set out in full under "Assumptions" in `?irt_causal_bart`; the three worth
stating up front:

* **The items must be measured before treatment.** If the responses postdate
  `z`, `theta` is a mediator or a collider rather than a confounder, and
  adjusting for it can move the answer in either direction.
* **`theta` is assumed to be the entire confounder.** There is no covariate
  argument, so anything not routed through the latent trait is unadjusted.
* **The measurement model does the identifying**, so misspecifying it is not
  benign. Fewer or noisier items attenuate the estimate toward zero, and
  residual confounding from a mismeasured adjustment variable is not guaranteed
  to be conservative in direction.

A comparator worth running: put the sum score, or the raw responses, straight
into the outcome surface. That needs weaker assumptions than the 2PL, and if the
two disagree the measurement model is doing the work and deserves scrutiny.

## Simulate data

`simulate_irt_causal()` draws from this model. Treatment depends on `theta`, so
a naive treatment/outcome comparison is confounded.

```{r simulate}
library(bairrtt)

sim <- simulate_irt_causal(n_persons = 300, n_items = 40, ate = -0.2, seed = 42)
str(sim, max.level = 1)

# naive (confounded) difference in means -- biased for the true effect (-0.2)
mean(sim$y[sim$z == 1]) - mean(sim$y[sim$z == 0])
```

## Fit

`irt_causal_bart()` runs the whole sampler. The run below is deliberately small
so the vignette builds quickly; recovery sharpens with more persons/items and
longer chains.

```{r fit}
fit <- irt_causal_bart(
  sim$responses, sim$y, sim$z,
  n_burnin = 100, n_sampling = 250, seed = 42
)
fit
```

`n_chains` defaults to four, so this is four independent chains; `n_cores`
(default `getOption("mc.cores")`) controls how many run at once without
affecting the draws. The chain axis is always the second one, so `fit$ate` is a
`250 x 4` matrix that `mean()`, `quantile()`, and `hist()` treat as the 1000
pooled draws it is.

## Treatment effect

The estimand is `E[f(1, theta) - f(0, theta)]`, accumulated from the response
surface each scan. Its posterior recovers the true effect (`-0.2`), unlike the
naive comparison above.

```{r ate}
c(mean = mean(fit$ate),
  lower = quantile(fit$ate, 0.025, names = FALSE),
  upper = quantile(fit$ate, 0.975, names = FALSE))

hist(fit$ate, breaks = 30, main = "Posterior of the treatment effect",
     xlab = "ATE")
abline(v = sim$ate, col = "red", lwd = 2)
```

## Item parameters

The IRT item parameters are recovered too. Difficulty is read on the observable
scale --- each item's expected proportion-correct in the population, its
*endorsement* probability --- rather than as the raw latent `beta`, whose scale
is fixed only by convention and is stretched by the finite number of items
measuring each person; discrimination sharpens as that number grows.

```{r items}
alpha_draws <- extract(fit, "alpha")   # pooled draws x items
beta_draws  <- extract(fit, "beta")
alpha_hat <- colMeans(alpha_draws)

# standard-normal population, for integrating each item's characteristic curve
grid <- qnorm(ppoints(200))

# an item's expected proportion-correct over the population: an observable-scale
# difficulty, free of the latent metric that inflates the raw beta coordinate
endorsement <- function(alpha, beta)
  vapply(seq_along(beta),
         function(i) mean(plogis(alpha[i] * (grid - beta[i]))),
         numeric(1))

pi_true <- endorsement(sim$alpha, sim$beta)
pi_hat  <- rowMeans(vapply(seq_len(nrow(beta_draws)),
                           function(s) endorsement(alpha_draws[s, ], beta_draws[s, ]),
                           numeric(ncol(beta_draws))))

c(alpha_cor      = cor(alpha_hat, sim$alpha),
  difficulty_cor = cor(pi_hat,    pi_true))

plot(qlogis(pi_true), qlogis(pi_hat),
     xlab = "true difficulty (logit endorsement)", ylab = "posterior mean",
     main = "Item difficulties"); abline(0, 1, col = "red")
```

## Diagnostics

The per-person `theta` acceptance rate is adapted toward its target (0.44 by
default) during burn-in and then frozen, and the WALNUTS item sampler is frozen
for the sampling phase.

```{r diagnostics}
colMeans(tail(fit$theta_accept, 250))  # sampling-phase theta acceptance
fit$theta_sd                           # frozen proposal SD, per chain
fit$tuning[[1]]$frozen                 # WALNUTS frozen?
```

Those are all *within*-chain. The question they cannot answer is whether the
chain went to the right place at all --- which matters here, because IRT models
invite multimodality. Nothing in the likelihood distinguishes `theta` from
`-theta` except the sign constraint on the discriminations, and with several
traits the labels can permute too. A single chain that settled in the wrong mode
looks perfectly healthy from the inside. That is why `n_chains` defaults to four.

`summary()` reports rank-normalized split R-hat with bulk and tail effective
sample size (via [posterior](https://cran.r-project.org/package=posterior) when
it is installed). Each chain is split in half first, so a within-chain trend
registers as between-chain disagreement.

```{r summary}
summary(fit)
```

**Read R-hat on the estimand, not on the maximum over everything.** R-hat is
conventionally judged against 1.01, but the *maximum* of many R-hats has a null
distribution that is not centred at one. With a few hundred effective draws per
parameter, a perfectly converged fit exceeds 1.01 on at least one of a few
hundred item and person parameters essentially always --- so a "max R-hat"
across all of them would condemn every fit this package produces, and condemn
larger ones harder. `summary()` therefore defaults to `ate` and `sigma`. The
item and person parameters are available on request and are worth reading, but
read their effective sample sizes rather than thresholding their R-hats:

```{r summary_items}
head(summary(fit, vars = "alpha")$stats, 3)
```

Pool across chains with `extract()`, \pkg{dbarts}'s generic:

```{r pool}
ate <- extract(fit, "ate")          # 250 x 4 draws, pooled
length(ate)
c(mean = mean(ate), lower = quantile(ate, 0.025, names = FALSE),
  upper = quantile(ate, 0.975, names = FALSE))
```

For contrast, a run cut far too short. Its `ate` posterior is not obviously
wrong to look at; the diagnostics are what tell you not to trust it.

```{r notconverged}
short <- irt_causal_bart(
  sim$responses, sim$y, sim$z,
  n_burnin = 5, n_sampling = 25, warmup_start = 2, seed = 42
)
summary(short)
```

## Observed covariates

The latent trait is rarely the only confounder. `x` adds observed covariates,
and they enter in two places: both BART surfaces, so the adjustment set becomes
`(theta, x)`; and the measurement model, as a latent regression
`theta ~ N(x'gamma, 1)`.

The second is the one that gets forgotten. A conditioning model that omits
variables the analyst later relates to `theta` shrinks those relationships
toward zero --- Mislevy's (1991) result, and the reason large-scale assessments
condition their plausible values on everything downstream.

```{r covariates}
simx <- simulate_irt_causal(
  n_persons = 300, n_items = 40, ate = -0.2,
  n_covariates = 2, covariate_effect = c(1, -0.8), seed = 42
)

unadjusted <- irt_causal_bart(simx$responses, simx$y, simx$z,
                              n_burnin = 100, n_sampling = 250, seed = 42)
adjusted   <- irt_causal_bart(simx$responses, simx$y, simx$z, x = simx$x,
                              n_burnin = 100, n_sampling = 250, seed = 42)

interval <- function(f) quantile(as.vector(f$ate), c(0.025, 0.975), names = FALSE)
rbind(unadjusted = interval(unadjusted),
      adjusted   = interval(adjusted),
      truth      = rep(simx$ate, 2))
```

Here `x` confounds directly *and* shifts the trait, so ignoring it is
misspecified in both places at once and the unadjusted interval misses. The
latent-regression coefficients come back too:

```{r gamma}
colMeans(extract(adjusted, "gamma"))   # intercept, then one per covariate
simx$gamma
```

Read those as *conditioning* coefficients, not as structural effects of `x` on
the trait --- they describe the population the traits are being estimated
within, and are routinely over-interpreted.

## More than one latent trait

Several traits can confound the same treatment and outcome, each measured by its
own item bank. Pass `responses` as a *list* of response matrices, one per trait;
every item loads on exactly one trait, and the banks may differ in item count so
long as they cover the same persons in the same order. Conditional on the traits
the banks factorize exactly, so each gets its own WALNUTS sampler.

The traits are visited in a systematic scan --- trait `k` is proposed and
installed for every person before trait `k + 1` --- because the joint install
into both BART models takes one shared column at a time. That is an ordinary
Metropolis-within-Gibbs sweep; what it does not do is move a person's whole
trait vector as a single accept/reject unit.

```{r simulate2}
sim2 <- simulate_irt_causal(
  n_persons = 300, n_items = c(30, 40), n_traits = 2,
  ate = -0.2, prognostic = c(1.2, 0.8), seed = 42
)
dim(sim2$theta)                       # one column per trait
sapply(sim2$responses, dim)           # 300 x 30 and 300 x 40

fit2 <- irt_causal_bart(
  sim2$responses, sim2$y, sim2$z,
  n_burnin = 100, n_sampling = 250, seed = 42
)
```

The estimand is unchanged in form --- `E[f(1, theta) - f(0, theta)]`, now
conditioning on the whole trait matrix. The per-bank quantities come back as
lists, one entry per trait; everything else keeps its single-trait shape.

```{r ate2}
c(mean = mean(fit2$ate),
  lower = quantile(fit2$ate, 0.025, names = FALSE),
  upper = quantile(fit2$ate, 0.975, names = FALSE))

# item difficulties still track the truth within each bank
beta2 <- extract(fit2, "beta")     # one pooled draws x items matrix per bank
vapply(1:2, function(k) cor(colMeans(beta2[[k]]), sim2$beta[[k]]), numeric(1))
```

Only the disjoint-bank case is supported. Genuine multidimensional IRT, where
one item loads on several traits, is identified only up to a rotation of the
loadings and needs a constraint that this interface does not express.

## Reusing the WALNUTS item sampler

The item-parameter sampler is also exported on its own, for use inside other
models. Create it once, adapt with `irt_warmup()` while conditioning on a
(possibly changing) `theta`, `irt_freeze()`, then draw with `irt_draw()`.

```{r engine}
s <- irt_item_sampler(sim$responses, alpha = rep(1, 40), beta = rep(0, 40),
                      seed = 1)
for (i in 1:50) invisible(irt_warmup(s, sim$theta))   # adapt
irt_freeze(s)
draw <- irt_draw(s, sim$theta)                         # c(alpha, beta)
head(draw)
```
